Lab 3: The Linear Model in R: Answer key

### 2023-01-13 Lab 3  Fitting LMs/Lab3_LM.Rmd
Add files via upload

---
title: "Lab 3: The Linear Model in R"
author: "Eric Parajon"
output: pdf_document
---
Adapted  from code and text from Isabel Laterzo, Simon Hoellerbauer and Chelsea Estancona.

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, include = TRUE, eval=FALSE)

@@

#Installing and loading packages for use
pacman::p_load(xtable,stargazer,palmerpenguins,tidyverse)

@@

First, we'll simulate some data to use. This will be a very useful skill in the future, so pay attention to what we do here.

We first create our variables and then stipulate what our effects should be.

Let's set our seed to 123.

For our $\mathbf{x}_1$ variable, let's take 20 draws from a $\mathcal{N}(1.3, 2.3)$ (using the `R` parameterization, where the second parameter is the standard deviation).

Let's set $\beta_0$ to be 17, and $\beta_1$ to be 1.3. We also can't forget to simulate $\boldsymbol{\epsilon}$! Let's $\boldsymbol{\epsilon} \sim \mathcal{N}(0, 3.4)$.

@@

What would we do in order to simulate a linear relationship with the variables and effects above? Create the outcome variable `y`.
```{r, eval = FALSE}
y <- _____________________
```

What do $\mathbf{x}_1$ and $\mathbf{y}$ look like together? Use the `plot` function to plot a scatter plot of $\mathbf{x}_1$ against $\mathbf{y}$.

@@

Now that we have simulated explanatory variables and set up a 'true' relationship between our explanatory variable (x1) and outcome variable, we can use R's built-in functions to evaluate our linear model. For a linear model, we will want to use the `lm()` function. 
The basic syntax for this function is `lm(Y ~ X1 + X2)`.

Use `summary()` on the model object (`m1` in this case) in order to see some of the most interesting information about the fitted model.

@@

We can also combine what we learned about using the tidyverse with linear modeling!

As we dive into learning linear models in `R`, we're going to use some data from the `mtcars`. But first, lets use the `tidyverse` to clean it up a bit. First, take a look at the mtcars data (its readily available so you don't really need to load it in!).

@@

Great. Let's say we want to analyze the role that `disp` and `cyl` play in influencing `mpg`. However, we're only interested in observations where `cyl` is greater than 4. Let's use `dplyr` and the `tidyverse` to only keep observations where `cyl` is greater than 4. Then, only keep the three above variables in the new data set. Label this data set `new_data`.

Hint: the `dplyr` functions you are interested in here are `filter()` and `select()` - don't forget to call their help files if you need more information (e.g., `?filter()` in your console).
```{r, eval = FALSE}

@@

Now that we have our data ready, we can use `lm()`.  Fill in the blank in the chunk below where Y is `mpg`, X1 is `cyl`, and X2 is `disp`. Use `summary()` on the model object (`m2` in this case) in order to see some of the most interesting information about the fitted model.

Remember, our data is `new_data`.

@@

The `summary` function provides lots of valuable information about the model we've created. We can also extract different information if we need, for example, just the coefficients, or just the residuals. Use `View()` on `m2`. What does it look like `m2` is? What do you think we could do if we wanted to access the different objects within `m2`?

@@

We can also use indexing to get to parts of these vectors - the output from a model is a list object, which means we can use brackets with $ indexing.

@@

What if we wanted to include an interaction term? Try interacting `disp` and `cyl`. 
```{r, eval = FALSE}
m3 <- lm(mp ~ disp + cyl + _______, data=new_data) #use * for interaction- include both variables

@@

We don't actually have to write all of the variables two times; the `lm` function will include the variables involved in the interaction even if we don't specify them individually in the formula. Try this out below, creating model object `m4` and compare it to `m3`.
```{r, eval = FALSE}
#m4:
m4<-lm(_________)

@@

How might we present these results in a table? Both `xtable()` and `stargazer()` are functions that produce LaTeX code that could be copy-pasted into a .tex document or included in an Rmarkdown document. Try both of these on one of the model object `m4` we created above and see what happens. Note, these are VERY flexible and can be customized to make very nice tables, the ones we are generating here are just basic. 
```{r, eval=FALSE, echo=TRUE}
#These produce code for a .pdf that we can copy and paste
xtable(______)
stargazer(_______)

@@

When presenting our results, we might be largely concerned with prediction: if I have a model and provide it with 'new' data, what should I expect? How confident are we in that expectation? We can use the handy `predict()` function here, which works with almost any type of model object.

Check out `predict.lm` using ?, as that is the method that `predict` invokes when we feed it an lm object.

@@

Note that when using predict, it is very important that our new observations have variables that bear the same name as in the formula. Otherwise `predict` won't know what to do! Try out the following:
```{r, eval = FALSE}
#recreating new_dat_n_1
new_dat_n_1_a <- with(mtcars, data.frame(x_vec_n, 4))

@@

# Part 3: LM practice
Using the palmerpenguines dataset, create an OLS model using multiple predictors and then display the results in a publication quality table (using stargazer).
```{r}
library(palmerpenguins)
library(stargazer)

@@

#Create table

stargazer(____)
```

>>>>

---
title: 'Lab 3: The Linear Model in R'
author: "Eric Parajon"
output:
  html_document:
    df_print: paged
  pdf_document: default
---
Adapted  from code and text from Isabel Laterzo, Simon Hoellerbauer and Chelsea Estancona.

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, include = TRUE, echo=TRUE, eval=FALSE)

@@

#Installing and loading packages for use
pacman::p_load(xtable,stargazer,palmerpenguins,tidyverse,texreg)

@@

+ First, we'll simulate some data to use. This will be a very useful skill in the future, so pay attention to what we do here.

+ We first create our variables and then stipulate what our effects should be.

+ Let's set our seed to 123.

+ For our $\mathbf{x}_1$ variable, let's take 20 draws from a $\mathcal{N}(1.3, 2.3)$ (using the `R` parameterization, where the second parameter is the standard deviation).

+ Let's set $\beta_0$ to be 17, and $\beta_1$ to be 1.3. We also can't forget to simulate $\boldsymbol{\epsilon}$! Let's $\boldsymbol{\epsilon} \sim \mathcal{N}(0, 3.4)$.

@@

+ What would we do in order to simulate a linear relationship with the variables and effects above? Create the outcome variable `y`.
```{r, eval = FALSE}
y <- _____________________
```

+ What do $\mathbf{x}_1$ and $\mathbf{y}$ look like together? Use the `plot` function to plot a scatter plot of $\mathbf{x}_1$ against $\mathbf{y}$.

@@

+ Now that we have simulated explanatory variables and set up a 'true' relationship between our explanatory variable (x1) and outcome variable, we can use R's built-in functions to evaluate our linear model. For a linear model, we will want to use the `lm()` function. 
The basic syntax for this function is `lm(Y ~ X1 + X2)`.

+ Use `summary()` on the model object (`m1` in this case) in order to see some of the most interesting information about the fitted model.

@@

+ We can also combine what we learned about using the tidyverse with linear modeling!

+ As we dive into learning linear models in `R`, we're going to use some data from the `mtcars`. But first, lets use the `tidyverse` to clean it up a bit. First, take a look at the mtcars data (its readily available so you don't really need to load it in!).

@@

+ Great. Let's say we want to analyze the role that `disp` and `cyl` play in influencing `mpg`. However, we're only interested in observations where `cyl` is greater than 4. Let's use `dplyr` and the `tidyverse` to only keep observations where `cyl` is greater than 4. Then, only keep the three above variables in the new data set. Label this data set `new_data`.

+ Hint: the `dplyr` functions you are interested in here are `filter()` and `select()` - don't forget to call their help files if you need more information (e.g., `?filter()` in your console).
```{r, eval = FALSE}

@@

+ Now that we have our data ready, we can use `lm()`.  Fill in the blank in the chunk below where Y is `mpg`, X1 is `cyl`, and X2 is `disp`. Use `summary()` on the model object (`m2` in this case) in order to see some of the most interesting information about the fitted model.

+ Remember, our data is called `new_data`.

@@

+ The `summary` function provides lots of valuable information about the model we've created. We can also extract different information if we need, for example, just the coefficients, or just the residuals. Use `View()` on `m2`. What does it look like `m2` is? What do you think we could do if we wanted to access the different objects within `m2`?

@@

+ We can also use indexing to get to parts of these vectors - the output from a model is a list object, which means we can use brackets with $ indexing.

@@

+ What if we wanted to include an interaction term? Try interacting `disp` and `cyl`. 
```{r, eval = FALSE}
m3 <- lm(mp ~ disp + cyl + _______, data=new_data) #use * for interaction- include both variables

@@

+ We don't actually have to write all of the variables two times; the `lm` function will include the variables involved in the interaction even if we don't specify them individually in the formula. Try this out below, creating model object `m4` and compare it to `m3`.
```{r, eval = FALSE}
#m4:
m4<-lm(_________)

@@

+ How might we present these results in a table? Both `xtable()`, `stargazer()`, and `texreg()` are packages that produce LaTeX code that could be copy-pasted into a .tex document or included in an Rmarkdown document. Try these on one of the model object `m4` we created above and see what happens. Note, these are VERY flexible and can be customized to make very nice tables, the ones we are generating here are just basic. 
```{r, eval=FALSE, echo=TRUE}
#These produce code for a .pdf that we can copy and paste
xtable(______)
stargazer(_______)
texreg(_______)

@@

2. When presenting our results, we might be largely concerned with prediction: if I have a model and provide it with 'new' data, what should I expect? How confident are we in that expectation? We can use the handy `predict()` function here, which works with almost any type of model object.

+ Check out `predict.lm` using ?, as that is the method that `predict` invokes when we feed it an lm object.

@@

+ Note that when using predict, it is very important that our new observations have variables that bear the same name as in the formula. Otherwise `predict` won't know what to do! Try out the following:
```{r, eval = FALSE}
#recreating new_dat_n_1
new_dat_n_1_a <- with(mtcars, data.frame(x_vec_n, 4))

@@

# Part 3 Exercise: LM practice

+ Using the palmerpenguines dataset, create a linear model model using multiple predictors of your choice and then display the results in a publication quality table (using stargazer).
```{r}
library(palmerpenguins)
library(stargazer)

@@

#Create nicely formatted table
stargazer(____)
```

### 2023-01-13 Lab 3  Fitting LMs/Lab3_LM_Answers.Rmd
Add files via upload

---
title: "Lab 3: The Linear Model in R"
author: "Eric Parajon"
output: pdf_document
---

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, include = FALSE, echo=TRUE)

@@

What would we do in order to simulate a linear relationship with the variables and effects above? Create the outcome variable `y`.
```{r}
y <- b0 + b1 * x1 + error
```

What do $\mathbf{x}_1$ and $\mathbf{y}$ look like together? Use the `plot` function to plot a scatter plot of $\mathbf{x}_1$ against $\mathbf{y}$.

@@

Now that we have simulated explanatory variables and set up a 'true' relationship between our explanatory variable (x1) and outcome variable, we can use R's built-in functions to evaluate our linear model. For a linear model, we will want to use the `lm()` function. The basic syntax for this function is `lm(Y ~ X1 + X2)`. Use `summary()` on the model object (`m1` in this case) in order to see some of the most interesting information about the fitted model.

@@

# Part 2: Combining LM and the tidyverse

```{r}
#examine data
summary(mtcars)
```

Great. Let's say we want to analyze the role that `disp` and `cyl` play in influencing `mpg`. However, we're only interested in observations where `cyl` is greater than 4. Let's use `dplyr` and the `tidyverse` to only keep observations where `cyl` is greater than 4. Then, only keep the three above variables in the new data set. Label this data set `new_data`.

Hint: the `dplyr` functions you are interested in here are `filter()` and `select()` - don't forget to call their help files if you need more information (e.g., `?filter()` in your console).

@@

Now that we have our data ready, we can use R's built-in functions to evaluate our linear model. For a linear model, we will want to use the `lm()` function. The basic syntax for this function is `lm(Y ~ X1 + X2)`. Fill in the blank in the chunk below where Y is `mpg`, X1 is `cyl`, and X2 is `disp`. Use `summary()` on the model object (`m1` in this case) in order to see some of the most interesting information about the fitted model.

Remember, our data is `new_data`.

@@

What if we wanted to include an interaction term?
```{r}
m3<-lm(mpg~disp+cyl+disp*cyl, data=new_data) #use * for interaction- include both variables

@@

We don't actually have to write all of the variables two times; the `lm` function will include the variables involved in the interaction even if we don't specify them individually in the formula. Try this out below, creating model object `m4` and compare it to `m3`.
```{r}
#m3:
m4 <- lm(mpg ~ disp*cyl, data=new_data)

@@

How might we present these results in a table? Both `xtable()` and `stargazer()` are functions that produce LaTeX code that could be copy-pasted into a .tex document or included in an Rmarkdown document. Try both of these on one of the model object `m4` we created above and see what happens. Note, these are VERY flexible and can be customized to make very nice tables, the ones we are generating here are just basic. 
```{r, eval=FALSE, echo=TRUE}

xtable(m4)
stargazer(m4)

#you can also add lots of useful information onto these
#check out guides online!
```

When presenting our results, we might be largely concerned with prediction: if I have a model and provide it with 'new' data, what should I expect? How confident are we in that expectation? We can use the handy `predict()` function here, which works with almost any type of model object. Check out `predict.lm` using ?, as that is the method that `predict` invokes when we feed it an lm object.

@@

# Part 3: LM practice
Using the palmerpenguines dataset, create an OLS using multiple predictors and then display the results in a publication quality table (using stargazer).
```{r}
library(palmerpenguins)
library(stargazer)

@@

#Putting results in table
stargazer(bill_model,title="Effect of Predictors on Bill Length",digits=3,star.cutoffs = c(0.1,0.05,0.01),
          label="bill_model",
          covariate.labels = c("\\textbf{Species} (reference= Adélie) &   \\\\
                      \\hspace{1cm}Chinstrap",
                      "\\hspace{1cm}Gentoo",
                    "\\textbf{Sex} (reference= Female) &   \\\\
                      \\hspace{1cm}Male",
                      "\\textbf{Body mass (grams)}"),
          notes="\\parbox[t]{12cm}{Coefficients reported from OLS regression models. Significance codes:*$<$ 0.10,**p$<$ 0.05, ***p$<$ 0.01, two-tailed tests.}",
          notes.append=FALSE,
          style="apsr",
          dep.var.caption = c(""),
          dep.var.labels.include = FALSE,
          table.layout ="-c-!t-s-n",
          header=FALSE,
          notes.align= "l",
          column.sep.width = "1pt",
          font.size = "scriptsize",
          no.space = TRUE)

>>>>

---
title: 'Lab 3: The Linear Model in R: Answer key'
author: "Eric Parajon"
output:
  html_document:
    df_print: paged
---

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, include = TRUE, echo=TRUE, eval=TRUE)

@@

+ What would we do in order to simulate a linear relationship with the variables and effects above? Create the outcome variable `y`.
```{r}
y <- b0 + b1 * x1 + error
```

+ What do $\mathbf{x}_1$ and $\mathbf{y}$ look like together? Use the `plot` function to plot a scatter plot of $\mathbf{x}_1$ against $\mathbf{y}$.

@@

+ Now that we have simulated explanatory variables and set up a 'true' relationship between our explanatory variable (x1) and outcome variable, we can use R's built-in functions to evaluate our linear model. For a linear model, we will want to use the `lm()` function. The basic syntax for this function is `lm(Y ~ X1 + X2)`. Use `summary()` on the model object (`m1` in this case) in order to see some of the most interesting information about the fitted model.

@@

# Part 2: Combining LM and the tidyverse
+ We can also combine what we learned about using the tidyverse with linear modeling!

+ As we dive into learning linear models in `R`, we're going to use some data from the `mtcars`. But first, lets use the `tidyverse` to clean it up a bit. First, take a look at the mtcars data (its readily available so you don't really need to load it in!).

```{r}
library(tidyverse)
#examine data
summary(mtcars)
```

+ Great. Let's say we want to analyze the role that `disp` and `cyl` play in influencing `mpg`. However, we're only interested in observations where `cyl` is greater than 4. Let's use `dplyr` and the `tidyverse` to only keep observations where `cyl` is greater than 4. Then, only keep the three above variables in the new data set. Label this data set `new_data`.

+ Hint: the `dplyr` functions you are interested in here are `filter()` and `select()` - don't forget to call their help files if you need more information (e.g., `?filter()` in your console).

@@

+ Now that we have our data ready, we can use `lm()`.  Fill in the blank in the chunk below where Y is `mpg`, X1 is `cyl`, and X2 is `disp`. Use `summary()` on the model object (`m2` in this case) in order to see some of the most interesting information about the fitted model.

+ Remember, our data is called `new_data`.

@@

+ What if we wanted to include an interaction term?
```{r}
m3<-lm(mpg~disp+cyl+disp*cyl, data=new_data) #use * for interaction- include both variables

@@

+ We don't actually have to write all of the variables two times; the `lm` function will include the variables involved in the interaction even if we don't specify them individually in the formula. Try this out below, creating model object `m4` and compare it to `m3`.
```{r}
#m3:
m4 <- lm(mpg ~ disp*cyl, data=new_data)

@@

+ How might we present these results in a table? Both `xtable()`, `stargazer()`, and `texreg()` are packages that produce LaTeX code that could be copy-pasted into a .tex document or included in an Rmarkdown document. Try these on one of the model object `m4` we created above and see what happens. Note, these are VERY flexible and can be customized to make very nice tables, the ones we are generating here are just basic. 
```{r, results='asis'}
#These produce code for a .pdf that we can copy and paste
xtable(m4)
stargazer(m4,type="html")
htmlreg(m4)

#you can also add lots of useful information onto these
#check out guides online!

+ When presenting our results, we might be largely concerned with prediction: if I have a model and provide it with 'new' data, what should I expect? How confident are we in that expectation? We can use the handy `predict()` function here, which works with almost any type of model object. Check out `predict.lm` using ?, as that is the method that `predict` invokes when we feed it an lm object.

@@

# Part 3 Exercise: LM practice

+ Using the palmerpenguines dataset, create a linear model model using multiple predictors of your choice and then display the results in a publication quality table (using stargazer).
```{r, results='hide'}
library(palmerpenguins)
library(stargazer)

@@

#Putting results in table (latex version)
stargazer(bill_model,title="Effect of Predictors on Bill Length",digits=3,star.cutoffs = c(0.1,0.05,0.01),
          label="bill_model",
          covariate.labels = c("\\textbf{Species} (reference= Adélie) &   \\\\
                      \\hspace{1cm}Chinstrap",
                      "\\hspace{1cm}Gentoo",
                    "\\textbf{Sex} (reference= Female) &   \\\\
                      \\hspace{1cm}Male",
                      "\\textbf{Body mass (grams)}"),
          notes="\\parbox[t]{12cm}{Coefficients reported from an OLS regression model. Significance codes:*$<$ 0.10,**p$<$ 0.05, ***p$<$ 0.01, two-tailed tests.}",
          notes.append=FALSE,
          style="apsr",
          dep.var.caption = c(""),
          dep.var.labels.include = FALSE,
          table.layout ="-c-!t-s-n",
          header=FALSE,
          notes.align= "l",
          column.sep.width = "1pt",
          font.size = "scriptsize",
          no.space = TRUE)
```

```{r, results='asis'}
#Putting results in table (html version)
stargazer(bill_model,title="Effect of Predictors on Bill Length",digits=3,star.cutoffs = c(0.1,0.05,0.01),
          label="bill_model",
          covariate.labels = c("\\textbf{Species} Chinstrap",
                      "\\textbf{Species} Gentoo",
                    "\\textbf{Male} (reference= Female)",
                      "\\textbf{Body mass (grams)}"),
          notes="\\parbox[t]{12cm}{Coefficients reported from an OLS regression model. Significance codes:*$<$ 0.10,**p$<$ 0.05, ***p$<$ 0.01, two-tailed tests. Species reference= Adélie}",
          notes.append=FALSE,
          style="apsr",
          dep.var.caption = c(""),
          dep.var.labels.include = FALSE,
          table.layout ="-c-!t-s-n",
          header=FALSE,
          notes.align= "l",
          column.sep.width = "1pt",
          font.size = "scriptsize",
          no.space = TRUE,
          type="html")
